Extracting Function Signatures from Program Traces

### 2020-05-14 paper/main.tex
fixes fixes fixes

R has one builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{types} lists all of the builtin types that are provided by the
language. They are the possible return values of \k{typeof}. There is no
intrinsic notion of subtyping in R. But, in many context a \k{logical} will
convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
off conversion can occur in corner cases, such as \k{1<"2"} holds and
\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
converted to an integer. R does not distinguish between scalars and vectors
(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
distinguished missing value denoted by \code{NA}. The default type of
\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.

@@

Scalar data types include the distinguished \k{NULL} value, which is also of
type \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types that are mostly not used by user code, they are listed in
Figure~\ref{types} for reference.

@@

%
%
\subsubsection{Overview}
The goal of our analysis is to discover argument types for functions,
builtins, and specials.  Naively intercepting function calls and recording
argument types is incorrect, as recall that R wraps arguments in promises
(Section~\ref{sec:R}).  To circumvent this obstacle, we instead
intercept the forcing of promises: when a promise is forced, we identify the
surrounding call, collect the type of the value, and add it to the call
trace associated with the current function call.  Once function execution
completes, we collect information about the return value, add it to the call
trace, and save the completed call trace to be written out later.
%
%
\subsubsection{Type Information}
Once a value is obtained, either on function return or when an argument
promise is forced, we probe the value for the following information:

@@

\item First, we wanted to ensure that \typetracer captured type for
  unevaluated arguments.  Not every argument of a function needs to be used
  during every function execution, and the approach we discussed above
  relies on argument promises being forced.  There are a few reasons for this.

@@

cars_model <- lm(dist ~ speed, data = cars)
cv.model(cars_model)
\end{lstlisting}

\caption{Struct example from the \code{agricolae} package.}
\label{fig:struct-ex}
\end{center}
\end{figure}

In the code snippet, the function \code{cv.model} takes an argument \code{x}, which we observed to always be a {\it linear model}, which is part of R's built-in statistics functionality.
Internally, linear models are represented as lists with named elements, and one may access these named elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.

In the code snippet, the function \code{generate.startweights} return a list with (at least) two named elements: \code{weights} and \code{exclude}, which are accessed through \code{result$weights} and \code{result$exclude} respectively.
We include a struct type in our language to capture this behaviour: we would encode the return type of \code{generate.startweights} as e.g. \code{struct<x: double[], y: logical[]>}.  
We also thought to encode data frames as structs, as they are essentially lists with named columns of homogeneous values, but it was difficult to draw any meaningful conclusions from the data as unfortunately structs introduced a lot of noise into our analysis.

It is common for R example code to use some built-in data sets, such as \code{mtcars}, a data set extracted from a Motor Trends US magazine in 1974.
These data sets are used to show off some package functionality with ``real data'', and when \typetracer is configured to detect structs we saw a marked increase in misrepresentative types.
We implemented a unification strategy for struct types, removing them if they co-occurred with list types, and turning them into lists if two co-occurring struct types shared no common names, but even so we found ourselves with function signatures that were full of noise introduced by the struct names.

>>>>

R has one builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{types} lists all of the builtin types that are provided by the
language. They are the possible return values of \k{typeof}. There is no
intrinsic notion of subtyping in R. But, in many context a \k{logical} will
convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
off conversions can occur in corner cases, such as \k{1<"2"} holds and
\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
converted to an integer. R does not distinguish between scalars and vectors
(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
distinguished missing value denoted by \code{NA}. The default type of
\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.

@@

Scalar data types include the distinguished \k{NULL} value, which is also of
type \k{NULL}, instance of classes written using the S4~\footnote{put footnote on S4 here} object system,
closures and environments.  The implementation of R has a number of other
types that are mostly not used by user code, they are listed in
Figure~\ref{types} for reference.

@@

%
%
\subsubsection{Overview}
The goal of our analysis is to discover argument types for functions,
builtins, and specials (R functions such as \{, \code{if}, etc.).  
First, recall that R wraps arguments in promises (Section~\ref{sec:R}).
Naively intercepting function calls and recording argument types would result in incomplete data:
While we can dig into promises without forcing them, if the innermost value is an unevaluated expression we are unable to probe it without evaluating it, which may alter program behaviour.
To circumvent this obstacle, we instead
intercept the forcing of promises: when a promise is forced, we identify the
surrounding call, collect the type of the value, and add it to the call
trace associated with the current function call.  Once function execution
completes, we collect information about the return value, add it to the call
trace, and save the completed call trace to be written out later.
%
%
\subsubsection{Type Information}
Once a value is obtained, either on function return or when an argument
promise is forced, we can probe it without interfering with program behaviour.
We query the value for the following information:

@@

\item

First, we wanted to ensure that \typetracer captured types for unevaluated arguments.  
Unfortunately, not every argument of a function is necessarily used during every function execution, meaning that some arguments may remain unevaluated (as they are never forced), and there are a few reasons for this.

@@

data(sweetpotato)
model<-aov(yield~virus, data=sweetpotato)
cv.model(model)
\end{lstlisting}

\caption{Struct example from a test in the \code{agricolae} package.}
\label{fig:struct-ex}
\end{center}
\end{figure}

In the code snippet, the function \code{cv.model} takes an argument \code{x}, which we observed to always be a {\it linear model}, which is part of R's built-in statistics functionality.
Internally, linear models are represented as lists with named elements, and one may access these named elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.
To describe these kinds of values, we included a struct type in our type system, allowing users to specify name, type pairs as part of list type definitions.

This example also illustrates an issue we grappled with when dealing with structs.
On line 8, we see the call \code{data(sweetpotato)}.
It is quite common for R example code to use some built-in data sets, such as \code{sweetpotato}, to show off some package functionality with ``real data''.
We found that the types of functions which use these built-in data sets (such as \code{aov}) became polluted with the names from the example sets.

% We also thought to encode data frames as structs, as they are essentially lists with named columns of homogeneous values, but it was difficult to draw any meaningful conclusions from the data as unfortunately structs introduced a lot of noise into our analysis.

When \typetracer is configured to detect structs we saw a marked increase in misrepresentative types.
We implemented a unification strategy for struct types, removing them if they co-occurred with list types, and turning them into lists if two co-occurring struct types shared no common names, but even so we found ourselves with function signatures that were full of noise introduced by the struct names.
